---
title: "Homework 3"
author: "Matthew T. Russell"
date: "10/20/2021"
output: github_document
---

```{r setup, load packages, include = FALSE}
library(tidyverse)
library(p8105.datasets)
library(lubridate)
library(knitr)
library(patchwork)
library(ggridges)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

###### EDA of Instacart data
* size and structure of the dataset
* description of key variables
* give illustrative examples of observations

```{r, load data in, message=FALSE}
data("instacart")


instacart <-
  instacart %>% 
  mutate(
    order_dow = order_dow + 1,
    order_dow = wday(order_dow, label = T)  
    )
```

```{r, plots eda, message=FALSE}
#bar plot: products ordered by dept

dept_less_100k_plot <- 
instacart %>% 
  group_by(department) %>% 
  summarize(
    n_obs = n()
  ) %>% 
  filter(department != "missing" & n_obs < 100000) %>% 
  ggplot(aes(x = reorder(department, n_obs), y = n_obs)) +
    geom_bar(stat = "identity") + 
    xlab("Department") + ylab("Count") +
    scale_y_continuous(breaks = c(0, 20000, 40000, 60000, 80000),
      labels = c("0", "20k", "40k", "60k", "80k")) +
    coord_flip() 

dept_ge_100k_plot <-
instacart %>% 
  group_by(department) %>% 
  summarize(
    n_obs = n()
  ) %>% 
  filter(department != "missing" & n_obs >= 100000) %>% 
  ggplot(aes(x = reorder(department, n_obs), y = n_obs)) +
    geom_bar(stat = "identity") + 
    xlab("Department") + ylab("Count") +
    scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000),
      labels = c("0", "100k", "200k", "300k", "400k")) +
    coord_flip()

(dept_ge_100k_plot + dept_less_100k_plot) +
  plot_annotation(
    title = 'Total products ordered by department'
    )

#number of orders by time of day and day of week

instacart %>% 
  select(order_id, order_dow, order_hour_of_day) %>% 
  distinct %>% 
  mutate(
    time_of_day =
      ifelse(order_hour_of_day %in% 5:11, "morning", 
        ifelse(order_hour_of_day %in% 12:16, "afternoon", 
          ifelse(order_hour_of_day %in% 17:20, "evening", "night")
        )
      ) %>% 
    factor(levels = c("morning", "afternoon", "evening", "night"))
    ) %>% 
  group_by(time_of_day, order_dow) %>% 
  summarize(
    n_orders = n()
  ) %>% 
  arrange(time_of_day) %>% 
  ggplot(aes(x = order_dow, y = n_orders , fill = time_of_day)) +
    geom_col() +
    xlab("Day of Week") + ylab("Order Count") +
    labs(fill = "Time of Day")


# instacart %>% 
#   select(order_number, days_since_prior_order) %>% 
#   distinct %>% 
#   ggplot(aes(x = order_number, y = days_since_prior_order)) +
#   geom_histogram(stat = "identity")

```

###### How many aisles are there, and which aisles are the most items ordered from?

```{r, message=FALSE}
aisles <-
  instacart %>% 
    count(aisle, name = "total_items_ordered", sort = T)
```

There are `r nrow(aisles)` aisles in the `instacart` data. The top five aisles ordered from are
`r head(pull(aisles, aisle), 5)`.


###### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r, message=FALSE}
aisles %>% 
  filter(total_items_ordered > 10000) %>% 
  ggplot(aes(x = reorder(aisle, total_items_ordered), y = total_items_ordered)) +
    geom_bar(stat = "identity", colour = "black", fill = "white") + 
    xlab("Aisle") + ylab("Count") +
    theme(axis.text.y = element_text(size = 7)) +
    coord_flip()
```

###### Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r, message=FALSE}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
  count(aisle, product_name, name = "qty_purchased") %>% 
  arrange(desc(qty_purchased)) %>% 
  group_by(aisle) %>% 
  slice(1:3) %>% 
  kable(caption = 
          "Three Most Popular Items in Baking Ingredients, Dog Food Care, and Packaged Fruits &
           Veggies Aisles")
```


###### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r, message=FALSE}
  instacart %>% 
    filter(product_name %in% c(
      "Pink Lady Apples", "Coffee Ice Cream"
    )) %>% 
    group_by(product_name, order_dow) %>% 
    summarize(mean_time = mean(order_hour_of_day)) %>%
    mutate(
      mean_time = (mean_time*3600) %>% 
      seconds_to_period() %>% 
      hms(), 
      mean_time = period(
        second = round(second(mean_time), 0), 
        minute = minute(mean_time), 
        hour = hour(mean_time)
        ), 
      mean_time = as.POSIXct(mean_time, format = "%H:%M:%S", origin = lubridate::origin, tz = "UTC"), 
      mean_time = format(mean_time, "%I:%M:%S %p")
      ) %>% 
    pivot_wider(
      names_from = "order_dow", 
      values_from = "mean_time"
    ) %>% 
    kable(caption = "Average Time Coffee Ice Cream & Pink Lady Apples are Ordered per Day of Week")
```



