---
title: "Homework 3"
author: "Matthew T. Russell"
date: "10/20/2021"
output: github_document
---

```{r setup, load packages, include = FALSE}
library(tidyverse)
library(p8105.datasets)
library(lubridate)
library(knitr)
library(patchwork)
library(ggridges)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

##### Load Instacart data

```{r, load data in, message=FALSE}
data("instacart")

instacart <-
  instacart %>% 
  mutate(
    order_dow = order_dow + 1,
    order_dow = wday(order_dow, label = T)  
    )
```

There are `r nrow(instacart)`observations and `r ncol(instacart)` in the `instacart` data where each observation represents a product that was purchased as part of an Instacart order. Key variables include:

*`reordered`: binary variable indicating if a user has ordered the product in the past  
*`order_dow`: the day of the week the order was placed  
*`order_hour_of_day`: the hour the order was placed  
*`days_since_prior_order`: the number of days since the user placed their last order with the maximum being 30 and NA if it's  the user's first order  

##### Illustrative examples

###### Bar plot: products ordered by dept

```{r, bar plot products dept, message=FALSE}
dept_products_total <-
  instacart %>% 
  group_by(department) %>% 
  summarize(
    n_obs = n()
  ) %>% 
  arrange(desc(n_obs)) %>% 
  filter(department != 'missing')

dept_less_100k_plot <- 
  dept_products_total %>% 
  filter(n_obs < 100000) %>% 
  ggplot(aes(x = reorder(department, n_obs), y = n_obs)) +
    geom_bar(stat = "identity") + 
    xlab("Department") + ylab("Count") +
    scale_y_continuous(breaks = c(0, 20000, 40000, 60000, 80000),
      labels = c("0", "20k", "40k", "60k", "80k")) +
    coord_flip() 

dept_ge_100k_plot <-
  dept_products_total %>% 
  filter(n_obs >= 100000) %>% 
  ggplot(aes(x = reorder(department, n_obs), y = n_obs)) +
    geom_bar(stat = "identity") + 
    xlab("Department") + ylab("Count") +
    scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000),
      labels = c("0", "100k", "200k", "300k", "400k")) +
    coord_flip()

(dept_ge_100k_plot + dept_less_100k_plot) +
  plot_annotation(
    title = 'Total products ordered by department'
    )
```

The five departments with the most purchased products are `r head(pull(dept_products_total, department), 5)`, and the five departments with the least purchased products are `r tail(pull(dept_products_total, department), 5)`.

###### Number of orders by time of day and day of week

```{r, col plot time of day and dow, message=FALSE}
order_time_by_day <-
  instacart %>% 
    select(order_id, order_dow, order_hour_of_day) %>% 
    distinct %>% 
    mutate(
      time_of_day =
        ifelse(order_hour_of_day %in% 5:11, "morning", 
          ifelse(order_hour_of_day %in% 12:16, "afternoon", 
            ifelse(order_hour_of_day %in% 17:20, "evening", "night")
          )
        ) %>% 
      factor(levels = c("morning", "afternoon", "evening", "night"))
      ) %>% 
    group_by(time_of_day, order_dow) %>% 
    summarize(
      n_orders = n()
    ) %>% 
    arrange(desc(n_orders))

total_orders_per_day <-
  order_time_by_day %>% 
  group_by(order_dow) %>% 
  summarize(
    total_orders = sum(n_orders)
  ) %>% 
  arrange(desc(total_orders))

order_time_by_day %>% 
  ggplot(aes(x = order_dow, y = n_orders , fill = time_of_day)) +
    geom_col() +
    xlab("Day of Week") + ylab("Order Count") +
    labs(fill = "Time of Day")

```

To find out how many orders were placed at each time of day, I considered 5 AM to 11 AM morning, 12 PM to 4 PM afternoon, 5 PM to 8 PM evening, and 9 PM to 4 AM night. The day with the most orders overall was `r head(pull(total_orders_per_day, order_dow), 1)`. with `r head(pull(total_orders_per_day, total_orders), 1)` orders, and the day with the least orders overall was `r tail(pull(total_orders_per_day, order_dow), 1)`. with `r tail(pull(total_orders_per_day, total_orders), 1)` orders. The most popular time to place an order was `r head(pull(order_time_by_day, order_dow), 1)`. `r head(pull(order_time_by_day, time_of_day), 1)` with `r head(pull(order_time_by_day, n_orders), 1)` orders placed, and the least popular time to place an order was `r tail(pull(order_time_by_day, order_dow), 1)`. `r tail(pull(order_time_by_day, time_of_day), 1)` with `r tail(pull(order_time_by_day, n_orders), 1)` orders placed.


#### How many aisles are there, and which aisles are the most items ordered from?

```{r, message=FALSE}
aisles <-
  instacart %>% 
    count(aisle, name = "total_items_ordered", sort = T)
```

There are `r nrow(aisles)` aisles in the `instacart` data. The top five aisles ordered from are
`r head(pull(aisles, aisle), 5)`.


##### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r, message=FALSE}
aisles %>% 
  filter(total_items_ordered > 10000) %>% 
  ggplot(aes(x = reorder(aisle, total_items_ordered), y = total_items_ordered)) +
    geom_bar(stat = "identity", colour = "black", fill = "white") + 
    xlab("Aisle") + ylab("Count") +
    theme(axis.text.y = element_text(size = 7)) +
    coord_flip()
```

##### Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r, message=FALSE}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
  count(aisle, product_name, name = "qty_purchased") %>% 
  arrange(desc(qty_purchased)) %>% 
  group_by(aisle) %>% 
  slice(1:3) %>% 
  kable(caption = 
          "Three Most Popular Items in Baking Ingredients, Dog Food Care, and Packaged Fruits &
           Veggies Aisles")
```

##### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week

```{r, message=FALSE}
  instacart %>% 
    filter(product_name %in% c(
      "Pink Lady Apples", "Coffee Ice Cream"
    )) %>% 
    group_by(product_name, order_dow) %>% 
    summarize(mean_time = mean(order_hour_of_day)) %>%
    mutate(
      mean_time = (mean_time*3600) %>% 
      seconds_to_period() %>% 
      hms(), 
      mean_time = period(
        second = round(second(mean_time), 0), 
        minute = minute(mean_time), 
        hour = hour(mean_time)
        ), 
      mean_time = as.POSIXct(mean_time, format = "%H:%M:%S", origin = lubridate::origin, tz = "UTC"), 
      mean_time = format(mean_time, "%I:%M:%S %p")
      ) %>% 
    pivot_wider(
      names_from = "order_dow", 
      values_from = "mean_time"
    ) %>% 
    kable(caption = "Average Time Coffee Ice Cream & Pink Lady Apples are Ordered per Day of Week")
```

# Problem 2

##### Load BRFSS and clean data

```{r, load and clean BRFSS data, message=FALSE}
data("brfss_smart2010")

brfss <-
  brfss_smart2010 %>% 
  janitor::clean_names() %>% 
  rename(
    state = locationabbr,
    county = locationdesc
  ) %>% 
  filter(
    topic == "Overall Health"
  ) %>% 
  select(!(
    c(class, topic, display_order, data_value_unit, data_value_type, 
      data_value_footnote_symbol, data_source, class_id, topic_id, location_id,
      question_id, respid, geo_location
    )
  )) %>%
  mutate(
    county = str_remove_all(county, "^\\w\\w - ") %>% 
      str_remove_all(" County$"), 
    response = factor(
      response, ordered = T, levels = c(
        "Poor", "Fair", "Good", "Very good", "Excellent"
      )
    )
  )
```

###### In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r, location 2002 and 2010, message=FALSE}
brfss %>% 
  filter(year == 2002) %>% 
  select(state, county) %>% 
  distinct() %>% 
  count(state) %>% 
  filter(n >= 7) %>% 
  arrange(desc(n)) %>% 
  kable(caption = "States observed 7 or more times, 2002")

brfss %>% 
  filter(year == 2010) %>% 
  select(state, county) %>% 
  distinct() %>% 
  count(state) %>% 
  filter(n >= 7) %>% 
  arrange(desc(n)) %>% 
  kable(caption = "States observed 7 or more times, 2010")
```

###### Spaghetti plot of average data_value over time among 'Excellent' responses:

```{r, spaghetti plot, message=FALSE, warning=FALSE}
brfss %>% 
  filter(response == "Excellent") %>% 
  select(year, state, data_value) %>% 
  group_by(year, state) %>% 
  summarize(mean_data = mean(data_value)) %>% 
  ggplot(aes(x = year, y = mean_data, group = state)) +
    geom_line() +
    xlab("Year") + ylab("Mean Data Value")
```






